Historical advances in medicine that did not need (or want) a randomized clinical

### 2020-01-25 paper/paper.Rmd
Reorganized rct not needed documents

---
title: "Historical advances in medicine that did not need (or want) a randomized clinical trial."
author: "Steve Simon"
output: html_document
bibliography: reference.bib
nocite: |
  @zika-overview, @zika-case-series
---

@@

Perhaps the most commonly cited discovery that did not use RCTs was the research that drew a causal link between cigarette smoking and cancer. It is still instructive, however, to review the series of studies that led to the 1964 Surgeon General's report on smoking and health. Although some evidence appeared earlier, the concern about smoking rose largely from five case-control studies, all published in 1950 (@report-anniversary). This research approach was very new, but the authors took great pains to control for recall bias and confounding (@statistics-in-action).

@@

>>>>

---
title: Historical advances in medicine that did not need (or want) a randomized clinical
  trial.
author: "Steve Simon"
nocite: |
  @zika-overview, @zika-case-series
output:
  word_document: default
  html_document: default
bibliography: reference.bib
---

@@

Perhaps the most commonly cited discovery that did not use RCTs was the research that drew a causal link between cigarette smoking and lung cancer. It is still instructive, however, to review the series of studies that led to the 1964 Surgeon General's report on smoking and health. Although some evidence appeared earlier, the concern about smoking rose largely from five case-control studies, all published in 1950 (@doll-hill-case-control , @levin-case-control, @mills-porter-case-control, @schrek-baker-case-control, @wynder-graham-case-control). This research approach was very new, but the authors took great pains to control for recall bias and confounding (@statistics-in-action).

@@

Around the same time, researchers set up prospective cohort studies to further investigate the link between smoking and lung cancer (@doll-hill-cohort, @dorn-cohort, @hammond-horn-cohort). These studies were massive (more than 460,000 patients total across the three studies) and followed these patients for multiple years. Needless to say, these studies took much longer to complete, but provided a key piece of additional evidence (@report-anniversary).

The researchers received a lot of criticism, of course, and not just from tobacco interests. R.A. Fisher, perhaps the most prominent statistician of the era, published a series of stinging critiques of the research (@fisher-pamphlet, @fisher-smoking-letter-1, @fisher-smoking-letter-2, @fisher-smoking-letter-3). It is difficult to parse these criticisms. One source (@fisher-error) attributes it a largely unsupportable explanation rooted in genetics combined with a focus on small details that failed to support a smoking-cancer link while ignoring the totality of the evidence.

The controversy inspired one of the prominent researchers in this area, Sir Austin Bradford Hill, to delineate his nine famous principles for establishing causation (@hill-criteria).
